---
title: "Data analysis with R and dplyr"
author: "themisbo"
date: "3/10/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data analysis of the billboard 100 data

This document is part of the showcase, where I replicate the same brief and simple analyses with different tools.

This particular file focuses on data analysis (a few queries) of the billboard 100 data from the tidytuesday project.

The data can be found in <https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-09-14>. They consist of two documents: *billboard.csv* contains information about the songs focusing on their position in the top100 list at different weeks. *audio_features.csv* contains information about specific attributes of the songs from spotify.

For the specific analysis I will use **R** and **dplyr** (plus **RMarkdown**).

We start by loading the packages:

```{r}
library(tidyverse) # metapackage for data analysis, including dplyr
library(lubridate) # package for date manipulation
```

and the datsets:

```{r}
billboard <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/billboard.csv')

audio_features <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/audio_features.csv')
```

We can have a look at the nature of the billboard data:

```{r}
glimpse(billboard)
```

and a few summary statistics:

```{r}
skimr::skim(billboard)
```

For the first main query, our aim is to select only the songs that have reached the No 1 spot of the billboard and see how many weeks they have stayed at the billboard in total:

```{r}
top_songs <- billboard %>%
  filter(peak_position == 1) %>%
  select(-url, -peak_position, - previous_week_position) %>%
  mutate(week_id = mdy(week_id)) %>%
  group_by(performer, song) %>%
  summarise(max_weeks = max(weeks_on_chart), .groups = 'drop') %>%
  arrange(desc(max_weeks))

top_songs %>%
  head(10)
```

We check the summary statistics of the audio features dataset:

```{r}
skimr::skim(audio_features)
```

For the second main query, our aim is to derive information about the peak position a song has reached in the billboard and the main spotify information. For this, we need to join the two datasets:

```{r}
data_join <- billboard %>%
  group_by(performer, song, song_id) %>%
  summarise(best_position = max(peak_position), .groups = 'drop') %>%
  inner_join(audio_features %>% select(song_id, spotify_genre, danceability, energy, key, loudness, speechiness, acousticness, instrumentalness, liveness, valence, tempo), by = "song_id") %>%
  arrange(performer)

data_join %>%
  head(10)
```

Note that many of the data from the audio_features table are empty, but this is not the matter at hand.